 

### 2021-01-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

**Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`).**
Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [@defrenneGlobalBufferingTemperatures2019], and recent evidence shows a similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a] and in the northwestern United States [@davis_microclimatic_2019]. Intact tropical forests have also been show to exhibit lower peak daytime $T_{air}$ than neighboring tree crop agricultural plantations (S. R. Hardwick et al., The relationship between leaf area index and microclimate in tropical forest and oil palm plantation: Forest disturbance drives changes in microclimate. Agric. Meteorol. 201, 187–195 (2015)). Even though canopy cover buffers maximum air temperatures in the understory, inevitable increases in understory temperatures continue to be observed as a result of global temperature increases. However, the buffering by canopy cover smooths this warming in the understory [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020;@rollinsonClimateSensitivityUnderstory]

@@

Thermal time constant, a leaf trait, quantifies heat flux in a leaf through incorporating heat stored versus heat exchanged with the environment, with relation to time. It comprises of several underlying traits -- leaf area, LMA, leaf width, leaf dry matter content and stomatal conductance [@michaletzEnergeticCarbonEconomic2016]. Typically, larger leaves with greater leaf width experience longer leaf cooling time and thus increased thermal time constant compared to leaves with smaller leaf width. Dissections in leaves, such as lobes, decrease leaf width and thus thermal time constant [@leighInfluenceLeafSize2017]. Additionally, a leaf with a smaller time constant experiences variable leaf temperatures in relation to the environment and exhibits instantaneous carbon assimilation rate and stable photosynthesis enabled by greater LMA [@michaletzEnergeticCarbonEconomic2016] Guessing, sun exposed leaves have lower thermal time constant than shaded?      
[I have Michaletz's code to calculate the thermal time constant, so if there are data, or general tendencies for changes in leaf area, LMA, leaf width and leaf dry matter content we can determine how these trends would translate into changes in thermal time constant. 
In practice, an important driver seems to be leaf water content per unit area (I think Lawren might have mentioned this as well in our meeting), so if we can conclude that that is something that scales with the vertical gradient, then we can make an informed guess about how the thermal time constant will change--Martijn Slot]

@@

High light stimulates the need for photoprotection in leaves in order to protect internal photosystem 2 from photoinhibition through D1 protein replacement. Thus leaves exposed to greater irradiance have greater pools of xanthophyll pigments expressed on a mass or area basis--photoprotective pigments comprising of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. These are found to be proportional to within-canopy light gradient at an inter- and intra-species level. Therefore, $VAZ$ is lower in shaded leaves and greater in sun leaves, as observed among temperate tree species [@niinemetsAcclimationHighIrradiance1998]. In neotropical forest, foliage exposed to higher irradiance in the canopy showed greater b-carotene and vioxanthin-cycle pool on an area or chl basis compared to low light leaves that showed greater mass based chlorophyll, lutein and neoxanthin carotenoids [Mastubara et al. 2009]

@@

In a tropical forest of Puerto Rico and temperate forests in Michigan, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]

@@

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil Forests despite much lower air temperature due to high solar radiation loads. However, leaf temperatures themselves were moderated by air temperature. [@fauset_differences_2018].
-	(Slot et al. 2019) and refs therein
-	(Fauset et al. 2018)
Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016). There's some hint of this in Bolstad et al. 1999.

@@

@@

**Isoprene production**
- high emission rates occur in the mid-canopy, and even from very small statured trees and shrubs, at a site in the Brazilian Amazon (Taylor et al., in prep).
- The high temperature carbon compensation point of sun leaves is higher in isoprene emitting species than in non-emitting species, so it is to be expected that other aspects of heat tolerance will be similarly affected by isoprenes (https://doi.org/10.1111/pce.13564).
-  Within species, isporene production scales with light/ T

>>>>

**Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`).**
Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [@defrenneGlobalBufferingTemperatures2019], and recent evidence shows a similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a] and in the northwestern United States [@davis_microclimatic_2019]. Intact tropical forests have also been show to exhibit lower peak daytime $T_{air}$ than neighboring tree crop agricultural plantations (S. R. Hardwick et al., The relationship between leaf area index and microclimate in tropical forest and oil palm plantation: Forest disturbance drives changes in microclimate. Agric. Meteorol. 201, 187–195 (2015)). Even though canopy cover buffers maximum air temperatures in the understory, inevitable increases in understory temperatures continue to be observed as a result of global temperature increases. However, the buffering by canopy cover smooths this warming in the understory, thereby exhibiting an understory temperature lag [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020;@rollinsonClimateSensitivityUnderstory]

@@

Thermal time constant, a leaf trait, quantifies heat flux in a leaf through incorporating heat stored versus heat exchanged with the environment, with relation to time. It comprises of several underlying traits -- leaf area, LMA, leaf width, leaf dry matter content and stomatal conductance [@michaletzEnergeticCarbonEconomic2016]. Typically, at a steady state, larger leaves with greater leaf width experience longer leaf cooling time and thus increased thermal time constant compared to leaves with smaller leaf width. Dissections in leaves, such as lobes, decrease leaf width and thus thermal time constant [@leighInfluenceLeafSize2017]. Additionally, a leaf with a smaller time constant experiences variable leaf temperatures in relation to the environment and exhibits instantaneous carbon assimilation rate and stable photosynthesis enabled by greater LMA [@michaletzEnergeticCarbonEconomic2016] Guessing, sun exposed leaves have lower thermal time constant than shaded?      
[I have Michaletz's code to calculate the thermal time constant, so if there are data, or general tendencies for changes in leaf area, LMA, leaf width and leaf dry matter content we can determine how these trends would translate into changes in thermal time constant. 
In practice, an important driver seems to be leaf water content per unit area (I think Lawren might have mentioned this as well in our meeting), so if we can conclude that that is something that scales with the vertical gradient, then we can make an informed guess about how the thermal time constant will change--Martijn Slot]

@@

High light stimulates the need for photoprotection in leaves in order to protect internal photosystem 2 from hronic photoinhibition through enabling D1 protein replacement. Thus leaves exposed to greater irradiance have greater pools of xanthophyll pigments expressed on a mass or area basis--photoprotective pigments comprising of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. These are found to be proportional to within-canopy light gradient at an inter- and intra-species level. Therefore, $VAZ$ is lower in shaded leaves and greater in sun leaves, as observed among temperate tree species [@niinemetsAcclimationHighIrradiance1998]. In neotropical forest, foliage exposed to higher irradiance in the canopy showed greater b-carotene and vioxanthin-cycle pool on an area or chl basis compared to low light leaves that showed greater mass based chlorophyll, lutein and neoxanthin carotenoids [Mastubara et al. 2009]

@@

In a tropical wet and moist forest of Puerto Rico and temperate forests in Michigan, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]

@@

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil Forests despite much lower air temperature due to high solar radiation loads. However, leaf temperatures themselves were moderated by air temperature. [@fauset_differences_2018].
- In three forest types-- temperate, tropical wet and tropical moist, $T_{leaf max}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in upper canopies of all sites, however $T_{leaf max}$ exceeded $T_{opt}$ at different heights and temperatures in all forests; with no significant relationship between $T_{opt}$ and height. Additionally, With increasing height, $T_{leaf} - T_{air}$ difference increased in the upper canopies of all forests, exhibiting greatest $T_{leaf} - T_{air}$ in temperate deciduous forest compared to tropical moist and wet.

-	(Slot et al. 2019) and refs therein
-	(Fauset et al. 2018)
Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016). There's some hint of this in Bolstad et al. 1999.

@@

T50 might not capture thermal stress or sensitivity to increasing temperature in foliage, because the degree at which T50 is observed in leaves is atan extreme temperature not usually experienced by leaves acclimated to different light environment, at a lower foliage thermal stress point  T50 might be irrelevant, but variation to extreme temperature response might be better understood through T50 [@perezPSIIHeatTolerances2020]

@@

**Isoprene production**
- high emission rates occur in the mid-canopy, and even from very small statured trees and shrubs, at a site in the Brazilian Amazon (Taylor et al., in prep).
- The high temperature carbon compensation point of sun leaves is higher in isoprene emitting species than in non-emitting species, so it is to be expected that other aspects of heat tolerance will be similarly affected by isoprenes (https://doi.org/10.1111/pce.13564).
-  Within species, isporene production scales with light/ T; however regarless of light, isoprene emission increases with T such as in understory shaded leaves?
